Reordering Elements in List CRDTs

### 2020-02-18 paper/list-move.tex
Add section on range moves

\usepackage{tikz}
\usetikzlibrary{arrows.meta}

@@

\author{Martin Kleppmann}
\email{mk428@cl.cam.ac.uk}
\orcid{0000-0001-7252-6958}
\affiliation{%
  \institution{Department of Computer Science and Technology}
  \streetaddress{15 JJ Thomson Avenue}
  \city{Cambridge}
  \state{}
  \postcode{CB3 0FD}
  \country{United Kingdom}
}

\begin{abstract}
    List CRDTs (Conflict-free Replicated Data Types) allow multiple users to concurrently insert and delete items in a shared list.
    However, existing algorithms behave poorly when users concurrently reorder items in the list, or move items to a new position.
    We describe an extension to list CRDTs that improves their support for reordering.
\end{abstract}

@@

Replicated lists can be used to implement a variety of important applications, such as:
\begin{itemize}
    \item text editors (text is a list of characters);
    \item graphics applications (a list of graphical objects, where the order determines visibility: objects ``further down'' may be occluded or filtered by the objects ``higher up'', as illustrated in Fig.~\ref{fig:smiley});
    \item to-do lists and task trackers (a list of tasks, where the order may reflect the relative priority of tasks as chosen by the user).
\end{itemize}

@@

All of the aforementioned algorithms allow replicas to \emph{insert} or \emph{delete} elements anywhere in the list.
However, none of them have explicit support for \emph{reordering} or \emph{moving} elements from one position to another position in the list.
This is a surprising omission because reordering is a commonly required operation: in many to-do list applications, a user can drag and drop list elements to reorder them, and graphics software allows users to reorder the object list with commands such as ``bring to front'' (which moves an object to the top of the list) and ``send to back'' (which moves an object to the bottom).

@@

\begin{figure*}
    \centering
    \vspace{3em}
    \begin{tikzpicture}[auto,scale=1.0]
        \tikzstyle{item}=[rectangle,draw,inner ysep=4pt,text height=9pt,text depth=2pt,minimum width=2.2cm]
        \tikzstyle{network}=[thick,dashed,blue,-{Stealth[length=3mm]}]
        \path [draw,dotted] (0,2) -- (17.5,2);
        \node at (1, 4) {Replica $A$:};
        \node at (1, 0) {Replica $B$:};
        \matrix [row sep={19pt,between origins},nodes=item] at (4, 4) {
            \node [minimum width=0.5cm] {1.}; & \node (top1) {buy milk}; \\
            \node [minimum width=0.5cm] {2.}; & \node {water plants}; \\
            \node [minimum width=0.5cm,fill=red!10] {3.}; & \node [fill=red!10] (joe1) {phone joe}; \\
        };
        \matrix [row sep={19pt,between origins},nodes=item] at (4, 0) {
            \node [minimum width=0.5cm] {1.}; & \node (top2) {buy milk}; \\
            \node [minimum width=0.5cm] {2.}; & \node {water plants}; \\
            \node [minimum width=0.5cm,fill=red!10] {3.}; & \node [fill=red!10] (joe2) {phone joe}; \\
        };
        \draw [->,red,thick] (joe1.east) to [in=0, out=0] node [right,near end] {move} (top1.north east);
        \draw [->,red,thick] (joe2.east) to [in=0, out=0] node [right,near end] {move} (top2.north east);
        \draw [->,thick] (6, 0) -- (8.5, 0);
        \draw [->,thick] (6, 4) -- (8.5, 4);
        \matrix [row sep={19pt,between origins},nodes=item] at (10, 4) {
            \node [minimum width=0.5cm,fill=red!10] {1.}; & \node [fill=red!10] {phone joe}; \\
            \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
        };
        \matrix [row sep={19pt,between origins},nodes=item] at (10, 0) {
            \node [minimum width=0.5cm,fill=red!10] {1.}; & \node [fill=red!10] {phone joe}; \\
            \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
        };
        \draw [->,thick] (11.5, 0) -- (14.5, 0);
        \draw [->,thick] (11.5, 4) -- (14.5, 4);
        \draw [network] (12.5, 0) to [out=90,in=270] (13.1, 4);
        \draw [network] (12.5, 4) to [out=270,in=90] (13.1, 0);
        \node [rotate=90,blue,font=\footnotesize] at (13.5, 2) {network communication};
        \matrix [row sep={19pt,between origins},nodes=item] at (16, 4) {
            \node [minimum width=0.5cm,fill=red!10] {1.}; & \node [fill=red!10] {phone joe}; \\
            \node [minimum width=0.5cm,fill=red!10] {2.}; & \node [fill=red!10] {phone joe}; \\
            \node [minimum width=0.5cm] {3.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {4.}; & \node {water plants}; \\
        };
        \matrix [row sep={19pt,between origins},nodes=item] at (16, 0) {
            \node [minimum width=0.5cm,fill=red!10] {1.}; & \node [fill=red!10] {phone joe}; \\
            \node [minimum width=0.5cm,fill=red!10] {2.}; & \node [fill=red!10] {phone joe}; \\
            \node [minimum width=0.5cm] {3.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {4.}; & \node {water plants}; \\
        };
    \end{tikzpicture}
    \caption{If reordering is implemented by deleting and re-inserting, concurrent moves of the same item result in the anomaly shown here: the moved item is duplicated.}
    \label{fig:duplication}
\end{figure*}

\begin{figure*}
    \centering
    \vspace{3em}
    \begin{tikzpicture}[auto,scale=1.0]
        \tikzstyle{item}=[rectangle,draw,inner ysep=4pt,text height=9pt,text depth=2pt,minimum width=2.2cm]
        \tikzstyle{network}=[thick,dashed,blue,-{Stealth[length=3mm]}]
        \path [draw,dotted] (0,1.5) -- (17.5,1.5);
        \node at (1, 3) {Replica $A$:};
        \node at (1, 0) {Replica $B$:};
        \matrix [row sep={19pt,between origins},nodes=item] at (4, 3) {
            \node [minimum width=0.5cm] {1.}; & \node (top1) {buy milk}; \\
            \node [minimum width=0.5cm] {2.}; & \node {water plants}; \\
            \node [minimum width=0.5cm,fill=red!10] {3.}; & \node [fill=red!10] (joe1) {phone joe}; \\
        };
        \matrix [row sep={19pt,between origins},nodes=item] at (4, 0) {
            \node [minimum width=0.5cm] {1.}; & \node (top2) {buy milk}; \\
            \node [minimum width=0.5cm] {2.}; & \node {water plants}; \\
            \node [minimum width=0.5cm,fill=red!10] {3.}; & \node [fill=red!10] (joe2) {phone joe}; \\
        };
        \draw [->,red,thick] (joe1.east) to [in=0, out=0] node [right,near end] {move} (top1.north east);
        \draw [->,red,thick] (joe2.east) to [in=0, out=0] node [right,at end,inner xsep=8pt] {move} (top2.south east);
        \draw [->,thick] (6, 0) -- (8.5, 0);
        \draw [->,thick] (6, 3) -- (8.5, 3);
        \matrix [row sep={19pt,between origins},nodes=item] at (10, 3) {
            \node [minimum width=0.5cm,fill=red!10] {1.}; & \node [fill=red!10] {phone joe}; \\
            \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
        };
        \matrix [row sep={19pt,between origins},nodes=item] at (10, 0) {
            \node [minimum width=0.5cm] {1.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm,fill=red!10] {2.}; & \node [fill=red!10] {phone joe}; \\
            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
        };
        \draw [->,thick] (11.5, 0) -- (14.5, 0);
        \draw [->,thick] (11.5, 3) -- (14.5, 3);
        \draw [network] (12.5, 0) to [out=90,in=270] (13.1, 3);
        \draw [network] (12.5, 3) to [out=270,in=90] (13.1, 0);
        \node [rotate=90,blue,font=\footnotesize] at (13.5, 1.5) {network communication};
        \matrix [row sep={19pt,between origins},nodes=item] at (16, 3) {
            \node [minimum width=0.5cm,fill=red!10] {1.}; & \node [fill=red!10] {phone joe}; \\
            \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
        };
        \matrix [row sep={19pt,between origins},nodes=item] at (16, 0) {
            \node [minimum width=0.5cm,fill=red!10] {1.}; & \node [fill=red!10] {phone joe}; \\
            \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
        };
    \end{tikzpicture}
    \caption{When the same item is concurrently moved to different positions on different replicas, one of those positions should become the ``winning'' position as the replicas converge. In this example, replica $A$'s move operation wins.}
    \label{fig:concurrent}
\end{figure*}

We argue that such duplication in the face of concurrency is undesirable, as it is generally not the behaviour expected by users.
For example, say a user has replicas of their to-do list on their laptop and their smartphone.
They reorder items on their laptop to match new priorities, and also reorder some items on their phone while the phone is offline, as illustrated in Fig.~\ref{fig:duplication}.
Later, when the phone is online again and synchronises with the laptop, any items that were moved on both devices will be duplicated.

This behaviour is especially confusing if the same item was moved to the same position (e.g.\ to the top of the list) on both devices, in which case the duplicated items are likely to be adjacent in the final list, as in Fig.~\ref{fig:duplication} (depending on any other insertions that might have taken place).
However, even if the concurrent move operations had different destination positions, duplicating the moved item at both destinations is unlikely to be desirable behaviour in most applications.

Rather, we argue that the best semantics in this situation is for the system to pick one of the destination positions as the ``winner''; as the replicas communicate, they all converge towards a list in which the moved element is located only at the winning position, and nowhere else.
This approach is illustrated in Fig.~\ref{fig:concurrent}.
Any deterministic method of picking the winner is suitable (e.g.\ based on a priority given to each replica, or based on the timestamp of the operations).

@@

Assume now that $p_1, p_2, \dots$ are position identifiers according to one of the aforementioned schemes.
We can now model the state of the list as a set of $(\mathit{value}, \mathit{register})$ pairs contained in an add-wins set CRDT:
\begin{align*}
\mathit{state} = \mathrm{AWSet}(\{ & (v_1, \mathrm{LWWRegister}(p_1)),\\
    & (v_2, \mathrm{LWWRegister}(p_2)), \;\dots\})
\end{align*}
In some schemes (such as Treedoc, Logoot and LSEQ), the position identifiers themselves encode their relative ordering~\cite{Interleaving}, and thus no further state is required.
In others (such as RGA and CT), the list CRDT requires an additional data structure that determines the order in which position identifiers appear in the list.

In order to iterate over the current state of the list, we need to query the above set by the position identifiers stored within the registers.
We can do this efficiently by using a database index, which allows us to find all the tuples within a set that contain a certain value (or range of values).

Note that when the position of a list element is changed, the old position identifier no longer has any value associated with it, and thus it should be regarded as nonexistent (like a tombstone in some CRDTs).
It is also possible for multiple list elements to have the same position identifier, and thus for several list elements to appear at the same position.
This situation is similar to having a multi-value register of values at each list position.
(Note this situation is different from our earlier use of a MV register, which was about one element having multiple positions, not one position having multiple elements.)

We can avoid having multiple list elements with the same position identifier by always creating a fresh position identifier as the destination of a move operation.
All of the aforementioned list CRDTs have methods of creating new position identifiers that are guaranteed to be globally unique.

>>>>

\usepackage[normalem]{ulem} % \sout{} for strikethrough text
\usepackage{tikz}
\usetikzlibrary{arrows.meta,snakes}

@@

\author{Martin Kleppmann}
\email{mk428@cl.cam.ac.uk}
\orcid{0000-0001-7252-6958}
\affiliation{%
  \institution{University of Cambridge}
  \streetaddress{15 JJ Thomson Avenue}
  \city{Cambridge}
  \state{}
  \postcode{CB3 0FD}
  \country{United Kingdom}
}

\begin{abstract}
Conflict-free Replicated Data Types (CRDTs) for lists allow multiple users to concurrently insert and delete elements in a shared list object.
However, existing algorithms behave poorly when users concurrently reorder elements in the list (i.e.\ move list elements to a new position).
We demonstrate the need for such a move operation, and describe an algorithm that extends a list CRDT with an explicit move operation.
Our algorithm can be used in conjunction with any existing list CRDT algorithm.
In addition to moving a single list element, we also discuss the open problem of moving ranges of elements.
\end{abstract}

\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10003752.10003809.10010172</concept_id>
       <concept_desc>Theory of computation~Distributed algorithms</concept_desc>
       <concept_significance>500</concept_significance>
   </concept>
   <concept>
       <concept_id>10003120.10003130.10003131.10003570</concept_id>
       <concept_desc>Human-centered computing~Computer supported cooperative work</concept_desc>
       <concept_significance>300</concept_significance>
   </concept>
   <concept>
       <concept_id>10002951.10003227.10003233</concept_id>
       <concept_desc>Information systems~Collaborative and social computing systems and tools</concept_desc>
       <concept_significance>300</concept_significance>
   </concept>
   <concept>
       <concept_id>10010520.10010521.10010537.10010540</concept_id>
       <concept_desc>Computer systems organization~Peer-to-peer architectures</concept_desc>
       <concept_significance>300</concept_significance>
   </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Distributed algorithms}
\ccsdesc[300]{Human-centered computing~Computer supported cooperative work}
\ccsdesc[300]{Information systems~Collaborative and social computing systems and tools}
\ccsdesc[300]{Computer systems organization~Peer-to-peer architectures}

\keywords{conflict-free replicated data types, distributed consistency, collaborative editing, optimistic replication}

@@

Replicated lists can be used to implement a variety of important applications, such as:
\begin{itemize}
    \item collaborative text editors (text is a list of characters);
    \item collaborative graphics applications (a list of graphical objects, where the order determines visibility: objects ``further down'' may be occluded or filtered by the objects ``higher up'', as illustrated in Figure~\ref{fig:smiley});
    \item to-do lists and task trackers (a list of tasks, where the order may reflect the relative priority of tasks as chosen by the user).
\end{itemize}

@@

All of the aforementioned algorithms allow replicas to \emph{insert} or \emph{delete} elements anywhere in the list.
However, none of them have explicit support for \emph{reordering} or \emph{moving} elements from one position to another position in the list.
This is a surprising omission because reordering is a commonly required operation: in many to-do list applications, a user can drag and drop list elements to reorder them, and graphics software allows users to reorder the object list with commands such as ``bring to front'' (which moves an object to the top of the list, so that it occludes other objects) and ``send to back'' (which moves an object to the bottom, so that it is occluded by other objects).

@@

\definecolor{insertcolor}{rgb}{0, 0.5, 0}
\definecolor{deletecolor}{rgb}{1, 0, 0}
\definecolor{movecolor}{rgb}{0.5, 0, 1}
\definecolor{networkcolor}{rgb}{0, 0.5, 1}

\begin{figure*}
    \centering
    \vspace{3em}
    \begin{tikzpicture}[auto,scale=1.0]
        \tikzstyle{item}=[rectangle,draw,inner ysep=4pt,text height=9pt,text depth=2pt,minimum width=2.2cm]
        \tikzstyle{network}=[thick,dashed,networkcolor,-{Stealth[length=3mm]}]
        \path [draw,dotted] (0,2) -- (17.5,2);
        \node at (1, 4) {Replica $A$:};
        \node at (1, 0) {Replica $B$:};
        \matrix [row sep={19pt,between origins},nodes=item] at (4, 4) {
            \node [minimum width=0.5cm] {1.}; & \node (top1) {buy milk}; \\
            \node [minimum width=0.5cm] {2.}; & \node {water plants}; \\
            \node [minimum width=0.5cm,fill=movecolor!10] {3.}; & \node [fill=movecolor!10] (joe1) {phone joe}; \\
        };
        \matrix [row sep={19pt,between origins},nodes=item] at (4, 0) {
            \node [minimum width=0.5cm] {1.}; & \node (top2) {buy milk}; \\
            \node [minimum width=0.5cm] {2.}; & \node {water plants}; \\
            \node [minimum width=0.5cm,fill=movecolor!10] {3.}; & \node [fill=movecolor!10] (joe2) {phone joe}; \\
        };
        \draw [->,movecolor,thick] (joe1.east) to [in=0, out=0] node [right,near end] {move} (top1.north east);
        \draw [->,movecolor,thick] (joe2.east) to [in=0, out=0] node [right,near end] {move} (top2.north east);
        \draw [->,thick] (6, 0) -- (8.5, 0);
        \draw [->,thick] (6, 4) -- (8.5, 4);
        \matrix [row sep={19pt,between origins},nodes=item] at (10, 4) {
            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
        };
        \matrix [row sep={19pt,between origins},nodes=item] at (10, 0) {
            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
        };
        \draw [->,thick] (11.5, 0) -- (14.5, 0);
        \draw [->,thick] (11.5, 4) -- (14.5, 4);
        \draw [network] (12.5, 0) to [out=90,in=270] (13.1, 4);
        \draw [network] (12.5, 4) to [out=270,in=90] (13.1, 0);
        \node [rotate=90,networkcolor,font=\footnotesize] at (13.5, 2) {network communication};
        \matrix [row sep={19pt,between origins},nodes=item] at (16, 4) {
            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm,fill=movecolor!10] {2.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm] {3.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {4.}; & \node {water plants}; \\
        };
        \matrix [row sep={19pt,between origins},nodes=item] at (16, 0) {
            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm,fill=movecolor!10] {2.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm] {3.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {4.}; & \node {water plants}; \\
        };
    \end{tikzpicture}
    \caption{If reordering is implemented by deleting and re-inserting, concurrent moves of the same element result in the anomaly shown here: the moved element is duplicated.}
    \label{fig:duplication}
\end{figure*}

\begin{figure*}
    \centering
    \vspace{3em}
    \begin{tikzpicture}[auto,scale=1.0]
        \tikzstyle{item}=[rectangle,draw,inner ysep=4pt,text height=9pt,text depth=2pt,minimum width=2.2cm]
        \tikzstyle{network}=[thick,dashed,networkcolor,-{Stealth[length=3mm]}]
        \path [draw,dotted] (0,1.5) -- (17.5,1.5);
        \node at (1, 3) {Replica $A$:};
        \node at (1, 0) {Replica $B$:};
        \matrix [row sep={19pt,between origins},nodes=item] at (4, 3) {
            \node [minimum width=0.5cm] {1.}; & \node (top1) {buy milk}; \\
            \node [minimum width=0.5cm] {2.}; & \node {water plants}; \\
            \node [minimum width=0.5cm,fill=movecolor!10] {3.}; & \node [fill=movecolor!10] (joe1) {phone joe}; \\
        };
        \matrix [row sep={19pt,between origins},nodes=item] at (4, 0) {
            \node [minimum width=0.5cm] {1.}; & \node (top2) {buy milk}; \\
            \node [minimum width=0.5cm] {2.}; & \node {water plants}; \\
            \node [minimum width=0.5cm,fill=movecolor!10] {3.}; & \node [fill=movecolor!10] (joe2) {phone joe}; \\
        };
        \draw [->,movecolor,thick] (joe1.east) to [in=0, out=0] node [right,near end] {move} (top1.north east);
        \draw [->,movecolor,thick] (joe2.east) to [in=0, out=0] node [right,at end,inner xsep=8pt] {move} (top2.south east);
        \draw [->,thick] (6, 0) -- (8.5, 0);
        \draw [->,thick] (6, 3) -- (8.5, 3);
        \matrix [row sep={19pt,between origins},nodes=item] at (10, 3) {
            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
        };
        \matrix [row sep={19pt,between origins},nodes=item] at (10, 0) {
            \node [minimum width=0.5cm] {1.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm,fill=movecolor!10] {2.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
        };
        \draw [->,thick] (11.5, 0) -- (14.5, 0);
        \draw [->,thick] (11.5, 3) -- (14.5, 3);
        \draw [network] (12.5, 0) to [out=90,in=270] (13.1, 3);
        \draw [network] (12.5, 3) to [out=270,in=90] (13.1, 0);
        \node [rotate=90,networkcolor,font=\footnotesize] at (13.5, 1.5) {network communication};
        \matrix [row sep={19pt,between origins},nodes=item] at (16, 3) {
            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
        };
        \matrix [row sep={19pt,between origins},nodes=item] at (16, 0) {
            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
        };
    \end{tikzpicture}
    \caption{When the same element is concurrently moved to different positions on different replicas, one of those positions should become the ``winning'' position as the replicas converge. In this example, replica $A$'s move operation wins.}
    \label{fig:concurrent}
\end{figure*}

We argue that such duplication in the face of concurrency is undesirable, as it is generally not the behaviour expected by users.
For example, say a user has replicas of their to-do list on their laptop and their smartphone.
They reorder items on their laptop to match new priorities, and also reorder some items on their phone while the phone is offline, as illustrated in Figure~\ref{fig:duplication}.
Later, when the phone is online again and synchronises with the laptop, any items that were moved on both devices will be duplicated.

This behaviour is especially confusing if the same item was moved to the same position (e.g.\ to the top of the list) on both devices, in which case the duplicated items are likely to be adjacent in the final list, as in Figure~\ref{fig:duplication} (depending on any other insertions that might have taken place).
However, even if the concurrent move operations had different destination positions, duplicating the moved element at both destinations is unlikely to be desirable behaviour in most applications.

Rather, we argue that the best semantics in this situation is for the system to pick one of the destination positions as the ``winner''; as the replicas communicate, they all converge towards a list in which the moved element is located only at the winning position, and nowhere else.
This approach is illustrated in Figure~\ref{fig:concurrent}.
Any deterministic method of picking the winner is suitable (e.g.\ based on a priority given to each replica, or based on the timestamp of the operations).

@@

Assume now that $p_1, p_2, \dots$ are position identifiers according to one of the aforementioned schemes, and $v_1, v_2, \dots$ are the values of list elements (e.g.\ to-do list items).
We can now model the state of the list as a set of $(\mathit{value}, \mathit{register})$ pairs contained in an add-wins set CRDT:
\begin{align*}
\mathit{state} = \mathrm{AWSet}(\{ & (v_1, \mathrm{LWWRegister}(p_1)),\\
    & (v_2, \mathrm{LWWRegister}(p_2)), \;\dots\})
\end{align*}
In some schemes (such as Treedoc, Logoot and LSEQ), the position identifiers themselves encode their relative ordering~\cite{Interleaving}, and thus no further state is required.
In others (such as RGA and CT), the list CRDT requires an additional data structure that determines the order in which position identifiers appear in the list.

In order to obtain the elements in their list order we need to query the above set by the position identifiers stored within the registers.
We can do this efficiently by treating the set of pairs as a table in a relational database and using a database index on the position-register column.
The index lets us efficiently find all the list elements with a particular position ID (or range of position IDs).

Note that when the position of a list element is changed, the old position identifier no longer has any value associated with it, and thus it should be regarded as nonexistent (like a tombstone in some CRDTs).
It is also possible for multiple list elements to have the same position identifier, and thus for several list elements to appear at the same position.
This is similar to having a multi-value register of values at each list position.
(Note this situation is different from our earlier use of a MV register, which was about one element having multiple positions, not one position having multiple elements.)

We can avoid having multiple list elements with the same position identifier by always creating a fresh position identifier as the destination of a move operation.
All of the aforementioned list CRDTs have methods of creating new, globally unique position identifiers for any location within a list.

This algorithm is a CRDT because its state is a straightforward composition of existing CRDTs.

\begin{figure*}
    \begin{tikzpicture}
        \tikzstyle{state}=[matrix,anchor=west,column sep={18pt,between origins},row sep={18pt,between origins}]
        \tikzstyle{char}=[draw,anchor=base,minimum width=18pt,text height=8pt,text depth=3pt,minimum height=18pt]
        \tikzstyle{mov}=[fill=movecolor!10]
        \tikzstyle{ins}=[fill=insertcolor!10]
        \tikzstyle{del}=[fill=deletecolor!10]
        \tikzstyle{network}=[thick,dashed,networkcolor,-{Stealth[length=3mm]}]
        \path [draw,dotted] (0,1.5) -- (17.5,1.5);
        \node at (0, 4) [anchor=west] {Replica $A$:};
        \node at (0, 1) [anchor=west] {Replica $B$:};
        \node (before1) at (0,3) [state] {
            \node [char] {$\bullet$}; &
            \node [char] {}; &
            \node [char] {\texttt{B}}; &
            \node [char] {\texttt{a}}; &
            \node [char] {\texttt{c}}; &
            \node [char] {\texttt{o}}; &
            \node [char] {\texttt{n}}; &
            \node [char] {\verb|\n|}; &
            \node {}; &
            \node {}; &
            \node {}; &
            \node {}; \\
            \node [char] {$\bullet$}; &
            \node [char] {}; &
            \node [char,ins] (ins-begin) {\color{insertcolor}{\texttt{S}}}; &
            \node [char,ins] {\color{insertcolor}{\texttt{o}}}; &
            \node [char,ins] {\color{insertcolor}{\texttt{y}}}; &
            \node [char,ins] {}; &
            \node [char,ins] (ins-end) {\color{insertcolor}{\texttt{m}}}; &
            \node [char,del] (del) {\color{deletecolor}{\renewcommand{\ULthickness}{1pt}\sout{~\texttt{M}~}}}; &
            \node [char] {\texttt{i}}; &
            \node [char] {\texttt{l}}; &
            \node [char] {\texttt{k}}; &
            \node [char] {\verb|\n|}; &
            \node {}; \\
        };
        \draw [snake=brace,raise snake=3pt,gap around snake=1pt,insertcolor,thick]
            (ins-end.south east) -- node [midway,below,inner ysep=8pt] {insert} (ins-begin.south west);
        \draw [snake=brace,raise snake=3pt,gap around snake=1pt,deletecolor,thick]
            (del.south east) -- node [midway,below,inner ysep=8pt] {delete} (del.south west);
        \node (before2) at (0,0) [state] {
            \node [char] {$\bullet$}; &
            \node [char] {}; &
            \node [char] {\texttt{B}}; &
            \node [char] {\texttt{a}}; &
            \node [char] {\texttt{c}}; &
            \node [char] {\texttt{o}}; &
            \node [char] {\texttt{n}}; &
            \node [char] (dest) {\verb|\n|}; \\
            \node [char,mov] {$\bullet$}; &
            \node [char,mov] {}; &
            \node [char,mov] {\texttt{M}}; &
            \node [char,mov] {\texttt{i}}; &
            \node [char,mov] {\texttt{l}}; &
            \node [char,mov] {\texttt{k}}; &
            \node [char,mov] (src) {\verb|\n|}; &
            \node {}; \\
        };
        \draw [->,movecolor,thick] (src.east) .. controls +(0:2cm) and +(0:1cm) .. node [right,near end,inner xsep=7pt] {move} (dest.north east);
        \draw [->,thick] (6.2, 0) -- (10.0, 0);
        \draw [->,thick] (8.0, 3) -- (10.0, 3);
        \draw [network] (8.5, 0) to [out=90,in=270] (9.1, 3);
        \draw [network] (8.5, 3) to [out=270,in=90] (9.1, 0);
        \node [rotate=90,networkcolor,font=\footnotesize] at (9.5, 1.5) {network communication};
        \node (after1) at (10,3) [state] {
            \node [char,mov] {$\bullet$}; &
            \node [char,mov] {}; &
            \node [char,mov] {\texttt{M}}; &
            \node [char,mov] {\texttt{i}}; &
            \node [char,mov] {\texttt{l}}; &
            \node [char,mov] {\texttt{k}}; &
            \node [char,mov] {\verb|\n|}; &
            \node {}; \\
            \node [char] {$\bullet$}; &
            \node [char] {}; &
            \node [char] {\texttt{B}}; &
            \node [char] {\texttt{a}}; &
            \node [char] {\texttt{c}}; &
            \node [char] {\texttt{o}}; &
            \node [char] {\texttt{n}}; &
            \node [char] {\verb|\n|}; \\
            \node [char,ins] {\color{insertcolor}{\texttt{S}}}; &
            \node [char,ins] {\color{insertcolor}{\texttt{o}}}; &
            \node [char,ins] {\color{insertcolor}{\texttt{y}}}; &
            \node [char,ins] {}; &
            \node [char,ins] {\color{insertcolor}{\texttt{m}}}; &
            \node {}; &
            \node {}; &
            \node {}; \\
        };
        \node (after2) at (10,0) [state] {
            \node [char,mov] {$\bullet$}; &
            \node [char,mov] {}; &
            \node [char,mov] {\texttt{M}}; &
            \node [char,mov] {\texttt{i}}; &
            \node [char,mov] {\texttt{l}}; &
            \node [char,mov] {\texttt{k}}; &
            \node [char,mov] {\verb|\n|}; &
            \node {}; \\
            \node [char] {$\bullet$}; &
            \node [char] {}; &
            \node [char] {\texttt{B}}; &
            \node [char] {\texttt{a}}; &
            \node [char] {\texttt{c}}; &
            \node [char] {\texttt{o}}; &
            \node [char] {\texttt{n}}; &
            \node [char] {\verb|\n|}; \\
            \node [char,ins] {\color{insertcolor}{\texttt{S}}}; &
            \node [char,ins] {\color{insertcolor}{\texttt{o}}}; &
            \node [char,ins] {\color{insertcolor}{\texttt{y}}}; &
            \node [char,ins] {}; &
            \node [char,ins] {\color{insertcolor}{\texttt{m}}}; &
            \node {}; &
            \node {}; &
            \node {}; \\
        };
    \end{tikzpicture}
    \caption{Replica $B$ moves a range of characters (here, ``$\bullet$ Milk''), while concurrently replica $A$ makes an edit inside the same range. If the move is performed character-by-character, $A$'s edit remains at the old position, while its surrounding context has moved to a new position, leading to an anomalous outcome.}
    \label{fig:range-bad}
\end{figure*}

\begin{figure*}
    \vspace{2em}
    \begin{tikzpicture}
        \tikzstyle{state}=[matrix,anchor=west,column sep={18pt,between origins},row sep={18pt,between origins}]
        \tikzstyle{char}=[draw,anchor=base,minimum width=18pt,text height=8pt,text depth=3pt,minimum height=18pt]
        \tikzstyle{mov}=[fill=movecolor!10]
        \tikzstyle{ins}=[fill=insertcolor!10]
        \tikzstyle{del}=[fill=deletecolor!10]
        \tikzstyle{network}=[thick,dashed,networkcolor,-{Stealth[length=3mm]}]
        \path [draw,dotted] (0,1.5) -- (17.5,1.5);
        \node at (0, 4) [anchor=west] {Replica $A$:};
        \node at (0, 1) [anchor=west] {Replica $B$:};
        \node (before1) at (0,3) [state] {
            \node [char] {$\bullet$}; &
            \node [char] {}; &
            \node [char] {\texttt{B}}; &
            \node [char] {\texttt{a}}; &
            \node [char] {\texttt{c}}; &
            \node [char] {\texttt{o}}; &
            \node [char] {\texttt{n}}; &
            \node [char] {\verb|\n|}; &
            \node {}; &
            \node {}; &
            \node {}; &
            \node {}; \\
            \node [char] {$\bullet$}; &
            \node [char] {}; &
            \node [char,ins] (ins-begin) {\color{insertcolor}{\texttt{S}}}; &
            \node [char,ins] {\color{insertcolor}{\texttt{o}}}; &
            \node [char,ins] {\color{insertcolor}{\texttt{y}}}; &
            \node [char,ins] {}; &
            \node [char,ins] (ins-end) {\color{insertcolor}{\texttt{m}}}; &
            \node [char,del] {\color{deletecolor}{\renewcommand{\ULthickness}{1pt}\sout{~\texttt{M}~}}}; &
            \node [char] {\texttt{i}}; &
            \node [char] {\texttt{l}}; &
            \node [char] {\texttt{k}}; &
            \node [char] {\verb|\n|}; &
            \node {}; \\
        };
        \draw [snake=brace,raise snake=3pt,gap around snake=1pt,insertcolor,thick]
            (ins-end.south east) -- node [midway,below,inner ysep=8pt] {insert} (ins-begin.south west);
        \draw [snake=brace,raise snake=3pt,gap around snake=1pt,deletecolor,thick]
            (del.south east) -- node [midway,below,inner ysep=8pt] {delete} (del.south west);
        \node (before2) at (0,0) [state] {
            \node [char] {$\bullet$}; &
            \node [char] {}; &
            \node [char] {\texttt{B}}; &
            \node [char] {\texttt{a}}; &
            \node [char] {\texttt{c}}; &
            \node [char] {\texttt{o}}; &
            \node [char] {\texttt{n}}; &
            \node [char] (dest) {\verb|\n|}; \\
            \node [char,mov] {$\bullet$}; &
            \node [char,mov] {}; &
            \node [char,mov] {\texttt{M}}; &
            \node [char,mov] {\texttt{i}}; &
            \node [char,mov] {\texttt{l}}; &
            \node [char,mov] {\texttt{k}}; &
            \node [char,mov] (src) {\verb|\n|}; &
            \node {}; \\
        };
        \draw [->,movecolor,thick] (src.east) .. controls +(0:2cm) and +(0:1cm) .. node [right,near end,inner xsep=7pt] {move} (dest.north east);
        \draw [->,thick] (6.2, 0) -- (10.0, 0);
        \draw [->,thick] (8.0, 3) -- (10.0, 3);
        \draw [network] (8.5, 0) to [out=90,in=270] (9.1, 3);
